\chapter{Ethics and Data Management}
\label{edm}

\section{Ethical reflection}

This thesis studies hybrid weather forecasting models using atmospheric reanalysis data
and numerical simulation. The project does not involve human participants, personal
data, or interventions on living subjects. The main ethical considerations are therefore
related to responsible modelling practice rather than privacy or informed consent. In
particular, machine-learning components in geophysical models can be difficult to
interpret, may behave non-physically outside the training distribution, and could be
over-trusted if evaluated only by aggregate forecast scores. For that reason, this thesis
does not treat predictive accuracy as the only objective: the learned closure is coupled
to a physics-based dynamical core, compared against numerical-only and machine-learning-only
baselines, and analysed with interpretable spectral-gain diagnostics. This helps make the
model behaviour more transparent and its limitations easier to assess. A further ethical
consideration is the computational and environmental cost of training large models. To
support responsible and reproducible research, the data sources, code, preprocessing
steps, and evaluation protocol should therefore be documented and archived as completely
as possible.

I acknowledge that the thesis adheres to the ethical code
(\url{https://student.uva.nl/en/information/ethics-in-research})
and research data management policies
(\url{https://rdm.uva.nl/en})
of UvA and IvI.

The following table lists the data used in this thesis (including source codes). I confirm
that the list is complete and the listed data are sufficient to reproduce the results of the
thesis. If a prohibitive non-disclosure agreement is in effect at the time of submission,
``NDA'' is written under ``Availability'' and ``License'' for the concerned data items.

\begin{table}[htbp]
\centering
\small
\begin{tabular}{|p{0.28\textwidth}|p{0.42\textwidth}|p{0.18\textwidth}|}
\hline
\textbf{\begin{tabular}[c]{@{}l@{}}Short description \\ (max. 10 words)\end{tabular}} &
\textbf{\begin{tabular}[c]{@{}l@{}}Availability \\ (e.g., URL, DOI)\end{tabular}} &
\textbf{\begin{tabular}[c]{@{}l@{}}License\end{tabular}} \\
\hline
ERA5 pressure-level reanalysis &
Copernicus Climate Data Store, DOI: 10.24381/cds.bd0915c6 &
CC-BY \\
\hline
ERA5 single-level reanalysis &
Copernicus Climate Data Store, DOI: 10.24381/cds.adbb2d47 &
CC-BY \\
\hline
Hybrid model source code &
\textbf{REPLACE WITH: GitHub URL or Zenodo DOI} &
MIT \\
\hline
\end{tabular}
\end{table}